---
title: "data.table"
author: "Tracey S. Frescino"
date: "Aug. 18, 2016"
output: pdf_document
---
<br>
<br>

## I am making this up


###Help Links
http://user2014.stat.ucla.edu/files/tutorial_Matt.pdf<br>
http://datatable.r-forge.r-project.org/datatable-intro.pdf

<br>

###Initialize R Environment<br>

```{r}

# Set working directory.
setwd("C:/_tsf/_GitHub/help/datatable")

# Set options bias for scientific notation
options(scipen=6)

# Load libraries
require(data.table)
require(microbenchmark)
require(dplyr)



# Read in Wyoming tree dataset to use as a bit data set to test with.
system.time(tree.df <- read.csv("data/WYtree.csv", header=TRUE))
class(tree.df)

# Converting a data frame to a data.table
# setDT converts lists and data.frames to data.tables by reference (no copy is made)
microbenchmark(
    tree.dt <- data.table(tree.df),
    tree.dt <- setDT(tree.df),
    tree.dt <- setDT(tree.df, key="PLT_CN")
)

## fread vs read.csv
## read.csv() reads rows into memory as character and then tries to convert them to integer and factors
## freed() reads everything as character

# Create a data.table from tree.df
system.time(tree.dt <- fread("data/WYtree.csv"))
class(tree.dt)
head(tree.dt)

# Create new data table to use as a small dataset to test with
## strlut
ESTUNIT <- c(1,1,3,3,5,7,7,9,9,11,11,13,13,15,17,19,19,21,23,23,25)
STRATA <-  c(1,2,1,2,2,1,2,1,2,1,2,1,2,2,2,1,2,2,1,2,2)
ACRES <- c(472611,2285002,245547,1776182,3072988,779515,4317444,215408,
           2514245,540035,1297089,935801,4994287,1428579,1283969,380750,
           2291052,1720074,762318,1854636,3440445)
NBRPLOTS <- round(runif(length(ESTUNIT), 1, 500))
strlut <- data.table(ESTUNIT, STRATA, ACRES)
strlut2 <- data.table(ESTUNIT, STRATA, NBRPLOTS)

unitvar <- "ESTUNIT"
strvar <- "STRATA"
acrevar <- "ACRES"


# Read in species look up table
ref_spcd <- fread("data/ref_spcd.csv")

```

<br>

###Meaning of data.table<br>

**DT[i, j, by]**

Take data.table **DT**, subset rows using **i**, then calculate **j** grouped by **by**

Relationship of commands to SQL

data.table    SQL
i             where
j             select
:=            update
by            group by
i             order by (in compound syntax)
i             having (in compound syntax)
nomatch=NA    outer join
nomatch=0     inner join

DT[where, select|update, group by][having][order by][ ]...[ ]


###Data Exploration 1 - subset columns<br>

```{r}
# Subset using numbers to identify columns
strlut[,1:2, with=FALSE]
#Bad:
#strlut[,1:2]

# Subset column as vector
strlut[,ESTUNIT]

# Subset column as data.table with 1 column
strlut[,list(ESTUNIT)]

# Subset column as vector, passing variable
strlut[[unitvar]]
strlut[,"ESTUNIT", with=FALSE]


# Subset 1 column as data.table, passing variable
strlut[,unitvar, with=FALSE]

# Subset more than 1 column as data.table, passing variable
strlut[,c(unitvar, strvar), with=FALSE]

# Subset rows 2 and 4 and columns ESTUNIT and STRATA
strlut[c(2,4), list(ESTUNIT, STRATA)]


# Subset rows 2 and 4 and columns ESTUNIT and STRATA, passing variables
STRATA <- "STRATA"
#strlut[c(2,4), list(get(eval(unitvar)), get(STRATA))]

strlut[c(2,4), list(get(eval(unitvar)), get(eval(STRATA)))]

# Subset a column and add another column that is product of 2 other columns
strlut[, list(ESTUNIT, newcol = ESTUNIT + STRATA)]

# Subset a column and add another column that is product of 2 other columns, passing vars 
strlut[, .(get(eval(unitvar)), newcol = get(unitvar) + get(strvar))]


## Get a count by ESTUNIT and STRATA
strlut[, table(ESTUNIT, STRATA)]
#makes a table

## or passing variables
strlut[, table(get(unitvar), get(strvar))]


## Add a column names STRWT that is proportion of ACRES by ESTUNIT
microbenchmark( strlut[,STRWT:=ACRES/sum(ACRES), by=ESTUNIT],
strlut.dplyr <-
  strlut %>%
  group_by(ESTUNIT) %>%
  mutate(STRWT = ACRES/sum(ACRES)))

strlut.dplyr <-
  strlut %>%
  group_by(ESTUNIT) %>%
  mutate(STRWT = ACRES/sum(ACRES))

## Passing variables
strlut[, STRWT:=get(eval(acrevar))/sum(get(eval(acrevar))), by=get(eval(unitvar))]

## Changing names of columns
setnames(strlut2, "NBRPLOTS", "n.strata")

## Subset unique values
subset(unique(strlut), select = unitvar)

## Get number of unique values
strlut[, uniqueN(get(unitvar))]


## Set column order
#setcolorder(strlut, c("ESTUNIT", "STRATA", "STRWT", "ACRES"))


```

<br>

<a name="sumgrp"></a>
###Data Exploration - Sum data by group<br>

```{r}

# Compare tapply and data.table method for speed
# Note: On small dataset, tapply is faster, but on larger dataset, the other is faster

tapply(strlut$ACRES, strlut$STRATA, sum)  ## data.frame method
strlut[, sum(ACRES), by=STRATA]           ## data.table method

# Using small dataset
microbenchmark(
  tapply(strlut$ACRES, strlut$STRATA, sum),
  strlut[, sum(ACRES), by=STRATA]
)

## Using larger dataset
microbenchmark(
  tapply(tree.df$BA, tree.df$PLT_CN, sum),
  tree.dt[, sum(BA), by=PLT_CN]
)


# Add a new column to data table by group (sum of basal area by species)
tree.dt[, sumba:=sum(BA, na.rm=TRUE), by=SPCD]
head(tree.dt)

# Remove a column from data table
tree.dt[,sumba:=NULL]
head(tree.dt)

# To pass a new column variable name in
newcol <- "sumba"
tree.dt[, (newcol):=sum(BA, na.rm=TRUE), by=SPCD]
head(tree.dt)


# Remove a column using a passed in variable 
tree.dt[, (newcol):=NULL]
head(tree.dt)


# Create a new data table with sum of basal area by species 
sumba <- tree.dt[, sum(BA, na.rm=TRUE), by=SPCD]
sumba

# Create a new data table (with new name) with sum of basal area by species 
spba <- tree.dt[, .(sumba=sum(BA, na.rm=TRUE)), by=SPCD]
spba

# Create a new data table (with new name) with sum of basal area by species - passing variable
ba <- "BA"
spba <- tree.dt[, list(sumba=sum(get(ba), na.rm=TRUE)), by=SPCD]
spba


# Sum basal area by 1 group (species=202)
tree.dt[SPCD==202, sum(BA, na.rm=TRUE)]


# Get a count of species that fall within height ranges (to nearest 10 ft) for
# height classes greater than 10.
#.N =freq
tree.dt[HT > 10, list(Count = .N), by = list(HTCL = 10 * round(HT / 10))]


## Calculate sum of BA by SPCD and HTCL (from above) 
tree.dt[HT > 10, .(sum(BA)), by = list(SPCD, HTCL = 10 * round(HT / 10))]

## Get sum of BA and number of trees by STATUSCD and by PLT_CN
tree.dt[PLT_CN==282479222489998, {
	sumba = sum(BA, na.rm=TRUE)
	n = .N
	.SD[, .(n, .N, sumba_in_STATUSCD = sum(BA, na.rm=TRUE), sumba_in_PLT=sumba), by=STATUSCD] }, 
	by=PLT_CN]



## Perform more than 1 operation on a column within a data.table
tree.dt[, {tmp1=BA*TPA_UNADJ; tmp2=mean(tmp1, na.rm=TRUE); tmp3=round(tmp2, 2)}, by=SPCD]

## Keep more than one variable
tree.dt[, {tmp1=BA*TPA_UNADJ; tmp2=mean(tmp1, na.rm=TRUE); tmp3=round(tmp2, 2);   list(tmp2=tmp2, tmp3=tmp3)}, by=SPCD]


## Multiply multiple columns by a constant
t1 <- copy(tree.dt)
t2 <- copy(tree.dt)
t3 <- copy(tree.dt)
t4 <- copy(tree.dt)
vars2convert <- c("CARBON_BG", "CARBON_AG")
microbenchmark(
    for(j in vars2convert){ set(t1, i=NULL, j=j, value=t1[[j]] * 1000) },
    t2[, (vars2convert) := lapply(.SD, function(x) x * 1000 ), .SDcols=vars2convert],
    t3[, (vars2convert) := lapply(.SD, "*", 1000 ), .SDcols=vars2convert],
    t4[, (vars2convert) := get(eval(vars2convert)) * 1000]
)
t1[1:2, c("PLT_CN", "TREE", vars2convert), with=FALSE]
t2[1:2, c("PLT_CN", "TREE", vars2convert), with=FALSE]
t3[1:2, c("PLT_CN", "TREE", vars2convert), with=FALSE]
t4[1:2, c("PLT_CN", "TREE", vars2convert), with=FALSE]


    ## CHANGE NA VALUES TO 0
#    for(col in tsumvarnmlst2) set(sumdat, which(is.na(sumdat[[col]])), col, 0) 


```



<br>

###Using Keys<br>

```{r}

# Set key for tree.dt as SPCD
setkey(tree.dt, SPCD)

# or if passing variable
var <- "SPCD"
setkeyv(tree.dt, var)


# Get sum basal area for spcd = 202
tree.dt[SPCD==202, sum(BA, na.rm=TRUE)]

# Get sum basal area for all species
tree.dt[, sum(BA, na.rm=TRUE), by=SPCD]

# Get sum of basal area, average height, maximum height, maximum diameter by species
# by= and keyby=  both retain row order within groups (by-order of first appearance)

key(tree.dt)
spsum <- tree.dt[,list(sumba=sum(BA, na.rm=TRUE),
              avght=mean(HT, na.rm=TRUE),
              maxht=max(HT, na.rm=TRUE),
              maxdia=max(DIA, na.rm=TRUE)), by=key(tree.dt)]
spsum

spsum2 <- tree.dt[,list(sumba=sum(BA, na.rm=TRUE),
              avght=mean(HT, na.rm=TRUE),
              maxht=max(HT, na.rm=TRUE),
              maxdia=max(DIA, na.rm=TRUE)), keyby=key(tree.dt)]
spsum2

# For just one species
# Note: Because key is numeric, must include list or J in front of category
sp202 <- tree.dt[list(202),list(sumba=sum(BA, na.rm=TRUE),
              avght=mean(HT, na.rm=TRUE),
              maxht=max(HT, na.rm=TRUE),
              maxdia=max(DIA, na.rm=TRUE)), keyby=key(tree.dt)]
sp202

# Without specifying key
sp202 <- tree.dt[list(202),list(sumba=sum(BA, na.rm=TRUE),
              avght=mean(HT, na.rm=TRUE),
              maxht=max(HT, na.rm=TRUE),
              maxdia=max(DIA, na.rm=TRUE))]
sp202


# For two species
microbenchmark(
sp202_746 <- tree.dt[list(c(202,746)),list(sumba=sum(BA, na.rm=TRUE),
              avght=mean(HT, na.rm=TRUE),
              maxht=max(HT, na.rm=TRUE),
              maxdia=max(DIA, na.rm=TRUE)), by=.EACHI],
sp202_746 <- tree.dt[list(c(202,746)),list(sumba=sum(BA, na.rm=TRUE),
              avght=mean(HT, na.rm=TRUE),
              maxht=max(HT, na.rm=TRUE),
              maxdia=max(DIA, na.rm=TRUE)), by=key(tree.dt)])
sp202_746


# All species
spall <- tree.dt[,list(sumba=sum(BA, na.rm=TRUE),
              avght=mean(HT, na.rm=TRUE),
              maxht=max(HT, na.rm=TRUE),
              maxdia=max(DIA, na.rm=TRUE)), by=key(tree.dt)]
spall


```

<br>

###Joining Tables<br>

Join type   DT syntax         Merge
INNER       X[Y, nomatch=0]   merge(X,Y,all=FALSE)
LEFT OUTER  Y[X]              merge(X,Y,all.x=TRUE)
RIGHT OUTER X[Y]              merge(X,Y,all.y=TRUE)
FULL OUTER  -                 merge(X,Y,all=TRUE)

```{r}

## Testing different ways of merging
setkey(tree.dt, SPCD)
ref_spcd <- setDT(ref_spcd, key="VALUE")

microbenchmark(
  a1 <- merge(tree.df, ref_spcd, by.x="SPCD", by.y="VALUE"), # using merge with data frame
  a2 <- merge(tree.dt, ref_spcd, by.x="SPCD", by.y="VALUE"), # using merge with data table
  a3 <- tree.dt[ref_spcd, on=c(SPCD="VALUE"), nomatch=0],    # using on with data table
  a4 <- tree.dt[ref_spcd, nomatch=0L],                       # using on with data table and keys
  setDT(tree.dt)[ref_spcd, on=c(SPCD="VALUE")],
  a5 <- inner_join(x=tree.df, y=ref_spcd, by = c("SPCD"="VALUE"))
  
)

## Testing group with merge as separate commands vs in same command
key(a4)
a4 <- tree.dt[ref_spcd, nomatch=0L]
a4[, sum(BA, na.rm=TRUE), by=key(a4)]

tree.dt[ref_spcd, nomatch=0, sum(BA, na.rm=TRUE), by=key(tree.dt)]
#or
merge.dt <- merge(tree.dt[,c("PLT_CN", "SPCD", "BA"), with=FALSE], ref_spcd, by.x="SPCD", by.y="VALUE")
head(merge.dt)

cols <- c("PLT_CN", "SPCD", "BA")
merge.dt <- merge(tree.dt[,cols, with=FALSE], ref_spcd, by.x="SPCD", by.y="VALUE")
head(merge.dt)

## Test difference between data.frame and data.table
# microbenchmark(
#   merge.df <- merge(tree.df[,c("PLT_CN", "SPCD", "BA")], ref_spcd, by.x="SPCD", by.y="VALUE"),
#   merge.dt <- merge(tree.dt[,c("PLT_CN", "SPCD", "BA"), with=FALSE], ref_spcd, by.x="SPCD", by.y="VALUE")
# )
  
microbenchmark(
  a4 <- tree.dt[ref_spcd, nomatch=0L],
  a4[, sum(BA, na.rm=TRUE), by=key(a4)],
  tree.dt[ref_spcd, nomatch=0, sum(BA, na.rm=TRUE), by=key(a4)]
)



## Merging a subset of one data table to another data table
a4 <- tree.dt[ref_spcd, nomatch=0L] 


## Add a new column to first data.table in merge using columns from second data.table
setkey(strlut, ESTUNIT, STRATA)
setkey(strlut2, ESTUNIT, STRATA)
strlut[strlut2, newcol:=ACRES*NBRPLOTS]


```

<br>

###Symbols (.N, .SD, .I, .BY, .GRP)<br>

```{r}

# Frequency table (Number of records by SPCD)
tree.dt[, .N, by=SPCD]

# Frequency table (Number of records by SPCD) - with named column
tree.dt[, .(NBR=.N), by=SPCD]

# Frequency table (Number of records by SPCD)
tree.dt[, as.data.table(table(SPCD))]


# Frequency table by 2 columns (Number of records by SPCD)
tree.dt[,.N, by=list(SPCD, STATUSCD)]
tree.dt[, as.data.table(table(SPCD, STATUSCD))]

## Passing variables
var1 <- "SPCD"
var2 <- "STATUSCD"
vars <- c(var1, var2)

tree.dt[,.N, by=c(var1, var2)]
tree.dt[,.N, by=vars]
tree.dt[,.(Freq=.N), by=vars]
tree.dt[SPCD %in% c(202, 746),.(Freq=.N), by=vars]


tree.dt[, as.data.table(table(get(var1), get(var2)))]


## Get PLT_CN values where there are more than 50 live trees
tuniqueid <- "PLT_CN"
tree.dt[STATUSCD == 1, (.N > 100), by=c(tuniqueid, "STATUSCD")][V1==TRUE][[tuniqueid]]

## Check results
tplt <- tree.dt[PLT_CN == 40405497010690]
dim(tplt)

tplt2 <- tree.dt[PLT_CN == 40406999010690]
dim(tplt2)

## Changing values of columns (ex. NA values to 0 values)
#na.to.0 <- function(x){x[is.na(x)] <- 0; x}
#sumtreef.prop[, (tdomscols) := lapply(.SD, na.to.0), .SDcols=tdomscols]

## this is faster
#for(col in tdomscols) set(sumtreef.prop2, which(is.na(sumtreef.prop2[[col]])), col, 0)
#DANGER ZONE

```

```{r}
mean.ht.dt <- tree.dt[,list(mean= mean(HT, na.rm=TRUE)), by="SPCD"]

mean.ht.dt2 <- tree.dt[,mean:= mean(HT, na.rm=TRUE), by="SPCD"]

library(ggplot2)
ggplot(mean.ht.dt, aes(y=mean, x=as.factor(SPCD))) + geom_bar(stat="identity")
```

